\AorE{Wavelet Scattering Transformasies Toegepas op Walvis Vokaliserings}{Wavelet Scattering Transforms Applied to Whale Vocalisations}

### 2024-05-27 dissertation/content/paper3/main.tex
started intro to p3
started intro
draft edit nearly complete, need to do refs

\chapter{Separable Wavelet Scattering}

\articleinfo{M.W. Rademan, D.J.J Verself, J.A. Du Preez}{IEEE Signal Processing Letters}{15 May 2024}{doi}{Wavelet scattering is a widely used feature extraction method due to it efficacy in extracting invariant features, 
while retaining any lost high frequency information resulting from averaging to obtain the desired amount of invariance. 
By generalising a 1-dimensional scattering transform, we extend its definition to an arbitrary number of independent dimensions. 
We show that, in a modern deep learning setting, separable wavelet scattering performs similarly to its non-separable counterparts 
with the MNIST hand-written digit dataset. We additionally demonstrate state-of-the-art results for a subset of the MedMNIST3D datasets.}

\section{Introduction}
\Ac{ws} has seen widespread use in classification applications as a powerful feature extraction method. It is an especially effective method for small datasets, since the feature filters are reminiscent of a \ac{cnn} front-end, while requiring no learning. The resulting features are invariant across all averaged dimensions, and exhibit separable class subspaces, allowing linear classifiers to be used with high efficacy \citep{2dscattering}.

Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \citep{1dscattering1, ws}, which was later extended to 2 \citep{2dscattering} and 3 dimensions \citep{3dscattering, harmonicscattering}. Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \citep{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \citep{ws_joint_tf, jointtfscattering2}. Joint-\ac{tf} scattering utilises a separable \ac{2d} filter that operates on the first level scattering scalogram, denoted by the operator $\mathcal{U}_1$, which is the only apparent usage of separable filters in the current scattering literature.

In a deep-learning setting, sensible initialisation of filters prior to learning can significantly improve performance and interpretability \citep{sincnet}. The greatest advantage of separable filters is their computation speed when computing convolutions directly, as is performed in \acp{cnn} \citep{separablecnn}. Additionally, separable filters reduce the number of parameters of a \ac{cnn} if the filters are configured to be learnable. Learnable filters can typically improve performance compared to their fixed-filters counterpart \citep{scattering_birdsong}.

At first glance, a \ac{ws} decomposition seems to be prohibitively expensive, but fast algorithms are possible due to the limited bandwidth of filter output. Fast algorithms utilise downsampling to take advantage of the demodulation of band-limited signals that result from the convolution of analytic wavelets \citep{2dscattering, 3dscattering, 1dscattering1}. Path pruning is also used to ignore filter combinations which have negligible energy. These algorithms are the standard implementation on many platforms, such as the Kymatio python package \citep{kymatio} and MATLAB \citep{MATLAB}.

@@

\subsection{Morlet}

We define a \ac{1d} zero-mean Gaussian as
\begin{equation}
    \theta_{\sigma_t}(t) = \frac{1}{\sqrt{2\pi\sigma_t^2}}e^{-\frac{1}{2}\left(\frac{t}{\sigma_t}\right)^2},
\end{equation}
which has a Fourier transform transform $\theta_{\sigma_t}(t) \xleftrightarrow{\mathcal{F}} \hat{\theta}_{\sigma_\omega}(\omega)$, where $\sigma_\omega = \frac{1}{\sigma_t}$.

For an arbitrary bandwidth definition, we define the bandwidth-to-$\sigma$ ratio $\beta \in \mathbb{R}^+$, such that the one-sided bandwidth of a zero-mean Gaussian is $\beta \sigma_\omega$. For reference, Kymatio's implementation of 2D scattering indirectly defines $\beta \approx 2.5$ \citep{kymatio}.

@@

A \ac{1d} Morlet $\uppsi$ has energy concentrated around 1 rad/s, with most of its energy contained in the interval $\omega \in [1 - \frac{1}{Q}, 1 + \frac{1}{Q}]$. $Q \in \mathbb{R}^+$ is defined as the number of wavelets per octave. The mother wavelet is given as
\begin{equation}
    \uppsi(t) = \theta_{\sigma_t}(t)\left(e^{jt} - \theta_0 \right),
\end{equation}
where $\theta_0 = \frac{\theta_{\sigma_t}(-1)}{\theta_{\sigma_t}(0)}$ to ensure zero mean: $\hat{\uppsi}(0) = 0$.

For reasons that will become apparent in section \ref{sec:filterbank}, we defined a 1D-wavelet dilated by a factor $\lambda$ as 
\begin{equation}
\label{eqn:dilwav}
    \uppsi_\lambda(t) = \begin{cases}
        \uppsi(\lambda t), \ \lambda \neq 0 \\
        \phi_j(t), \ \lambda = 0
    \end{cases},
\end{equation}
where $\phi$ represents the 1D \ac{lpf} utilised in scattering computations for the current dimension ($j$).

A $n$-dimensional separable wavelet may be constructed similarly, with
\begin{equation}
    \uppsi_{\vect{\lambda} }(\mathbf{u}) = \prod_{i=0}^{n-1} \uppsi_{\lambda_i}(u_i),
\end{equation}
where $\mathbf{u} = \left(u_1, ..., u_n\right)^T$ is the vector containing the dimensions of interest.

@@

\subsection{1D Filterbank Construction}
\label{sec:filterbank}
 Suppose a 1D Morlet has a centre frequency $\lambda_0$. The following filter is placed at $\lambda_0 2^{\frac{1}{Q}}$. To define the amount of overlap between filters, it is useful express the corresponding filter's frequency \ac{std} $\sigma_\omega$ in terms of the distance between filters in the frequency domain. If the wavelet at $\lambda_0$ requires a decay equivalent to $\alpha \sigma_\omega$ \ac{std}s at the next wavelet at $\lambda_0 2^{\frac{1}{Q}}$. It follows that $\sigma_\omega = \frac{1}{\sigma_t} = \frac{1}{\alpha}\left( 2^\frac{1}{Q} - 1 \right)$.

@@

The provided definitions of filter clearance ($\alpha$) and bandwidth ($\beta$) may also be extended to non-separable wavelets, such as the rotational construction employed in \citep{2dscattering}. However, a conversion is required to find the bandwidth in each axis for a specific wavelet for rotationally constructed filterbanks, in order to employ the downsampling strategies proposed in this paper.

@@

The scattering transform requires that the time support of all filters do not exceed the time support of $\phi$, ensuring no filters contain time information exceeding the invariance scale $d$. $d$ is equivalently defined as the total downsample factor of the scattering tranform, and is unique for each dimension. Equivalently, the frequency \ac{std} of the filters may not exceed the \ac{std} of $\phi(\omega)$. For brevity, we denote the frequency \ac{std}s as $\sigma_{\lambda, \omega}$ and $\sigma_{\phi, \omega}$ for the Morlets and the \ac{lpf} respectively.

@@

Only positive $\lambda$'s have been defined thus far, which provides inadequate coverage of the frequency domain in multiple dimensions. For real input signals, it is only necessary to cover half of one of the dimensions (only positive $\lambda$'s), whereas full coverage (both negative and positive $\lambda$'s) is required for additional dimensions. On-axis coverage is also required, in which each $\uppsi_\lambda$ must be multiplied with a Gaussian (zero-frequency wavelet), which the dilated wavelet definition in equation (\ref{eqn:dilwav}) defines as $\lambda=0$. A similar construction procedure is followed in \citep{jointtfscattering2}.

Given $m$ 1D filterbanks, with $m \ge 2$, each having Morlet filters with a positive set of lambdas $\Lambda_i^+$ and invariance scales $\vect{d} \in \mathbb{N}^m$, with $i = 1, ..., m$ indexing the dimension, we construct the $m$-dimensional filterbank with 
\begin{equation}
    \mathbb{F} = \left\{ \uppsi_{\vect{\lambda} }(\mathbf{u}) \ \middle| \ \vect{\lambda} \in \mathbb{L} \setminus \{\vect{0}\} \right\},
\end{equation}
where $\vect{u}$ is the $m$-dimensional spatial and/or time variable in which the each 1D filterbank is defined.  By definition, 
\begin{gather}
    \mathbb{L} = (\Lambda_1^+ \!\cup\! \{0\})\!\times\! (\Lambda_2^+ \!\cup\! \Lambda_2^- \!\cup \!\{0\}) \!\times\! ... \!\times\! (\Lambda_m^+ \!\cup\! \Lambda_m^- \!\cup\! \{0\}); \\
    \Lambda_i^- = \left\{-\lambda \ \middle| \ \lambda \in \Lambda_i^+\right\},
\end{gather}
where $\cup$ indicates the set union operator and $\times$ the Cartesian product. For $m=1$, the provided definitions result in a conventional \ac{1d} scattering transform \citep{1dscattering1}.

@@

% The dilated wavelet definition provided by (\ref{eqn:dilwav}) therefore provides a concise set construction, with which to construct the \ac{lpf} and on-axis wavelets simultaneously.

% All on-axis filters will be highly correlated with their negative counterpart due to the symmetry of the multidimensional Fourier transform of real signals. For example, the 2D filter specified by $\lambda = [0, x]^T$ will be highly correlated with the filter at $[0, -x]^T$. As the invariance scale $d$ increases, the on-axis filters will become more narrow in the zero coordinate dimension, thereby becoming increasingly correlated.

% \subsection{Discretisation}

% Morlets do not have compact support, as such, they must be truncated for computation. The invariance scale $d$ specifies the maximum allowable time support of all filters. Since input signals must be at least of length $d$, filters can be precomputed using the padded signal length, while also guaranteeing a suitable level of truncation. The padding required is on the order of $d$ samples.

% Note that, due to their separability, all $\uppsi_{\vect{\lambda}} \in \mathbb{F}$ can be stored as $d$ one-dimensional filters, and not as $N_1 \times ... \times N_d$ tensors, thereby saving on the storage required for pre-computed filters. The savings on storage space allow implementations to compute the filters at multiple input sample frequencies, allowing for optimal downsampling strategies without incurring a significant storage cost.

\section{Separable Scattering Transform} \label{sec:wst}
% In this section, we describe the efficient computation of $m$-dimensional separable scattering transforms. The described downsampling strategy is utilised to a limited extent in Kymatio scattering implementations \citep{kymatio} and the original proposed fast algorithms \citep{2dscattering}, which is generalised further in this section.

@@

A downsample factor $d$ which decomposes into a set of $n$ prime factors $\{p_1, ..., p_n\}$ with a corresponding multiset $\mathbb{M} = \left\{ m_1, ..., m_n \right\}$, where $m_i$ is the multiplicity of the prime $p_i$. We can find an optimal $d$ by maximising the sum $\sum\limits_{m \in M} m$.

@@

We can efficiently compute $s_1$ using compounded downsampling steps:
\begin{equation}
    \left(s_1\right)_{\downarrow d} = \left( \left(u_1\right)_{\downarrow d_1} * (\phi)_{\downarrow d_1}\right)_{\downarrow d_2},
\end{equation}
such that $d = d_1 \cdot d_2, \ d_1, d_2 \in \mathbb{N}^+$, with $d_1 | d$ and $d_1 \le  d_{\uppsi_1}$. In order to find $d_1$, we decrement $d_{\uppsi_1}$ until it divides $d$ evenly. Each scalogram $u_j[n]$ is not necessarily downsampled optimally, but has a downsampling factor which guarentees a consistent scattering coefficient output sample frequency.

@@

% Since the filterbank configuration is known, all required combinations of downsampled filters and their downsampling factors can be precomputed.

\begin{table*}[t!]

@@

\caption{MedMNIST classification results of compared to baseline \ac{nn} approaches (\citep{medmnist})} \label{tab:medmnist}

@@

@@

Given that $r|N$, we can express (\ref{eqn:circconv}) when downsampled as a periodised summation \citep{waveletsandsubbandcoding} in the frequency domain
\begin{equation}
    (x \otimes \uppsi [n])_{\downarrow r} \xleftrightarrow{\mathcal{FFT}}  \frac{1}{r} \sum_{i=0}^{r-1} \hat{x} \cdot \hat{\uppsi} [k + iN/r], %https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=5216ea733e562541b33a7f97dab0de072b2e8827
\end{equation}
which can be efficiently implemented via shape manipulation of tensors in computational packages like MATLAB or PyTorch \citep{pytorch}. To compute valid convolutions, we must pad $x$ and $\uppsi$ to have a total length of $N = N_x + d + c$, where $c \in \mathbb{N}^+$ is a constant that ensures that $d | N$.

% Note that only $d$ additional samples are required instead of $2d$, since $d$ is, by definition, the time support of $\phi$ in number of samples. Padding is applied equally to the left and right of $x$. For further reduction of boundary effects, reflection padding is utilised. Kymatio's 2D scattering implementation utilises a similar padding scheme.

% Thus far, the techniques discussed may be equivalently implemented on many scattering transforms. Separable scattering benefits from reduced computations resulting from filter separability. In particular, given a $m$-dimensional signal $x[\vect{n}]$ indexed by $\vect{n} = (n_1, ..., n_m)^T$, with each dimension having downsample factors expressed as a vector $\vect{r} = (r_1, ..., r_m)^T$, we can chain downsampling for each dimension:
% \begin{equation}
%     \left(x * \uppsi_{\vect{\lambda}}[\vect{n}]\right )_{\downarrow \vect{r}} = \left( \left(x[\vect{\lambda}] * \uppsi_{\lambda_i}[n_i]\right)_{\downarrow r_1} ... * \uppsi_{\lambda_m}[n_m]\right)_{\downarrow r_m}.
% \end{equation}.

% We abuse notation to indicate that $\downarrow r_i$ is an operator that downsamples the $i$'th dimension, and $\downarrow \vect{r}$ downsamples all dimensions each with their correponding downsample factor in $\vect{r} \in \mathbb{N}^m$.

% For $m$-dimensional sequences of dimensional lengths $N_1, ..., N_m$, the forward and inverse \acp{fft} computation time is reduced from $2\sum_{i} T \log N_i$ (no frequency periodisation), with $T = \prod_i N_i$ to $\sum_{i} \frac{T}{K_i} \log \frac{N_i}{K_i} + P \log P$, where $K_1 = 1$, $K_i = \prod_{j < i} r_j$ and $P = \prod_i \frac{N_i}{r_i}$. However, the number of convolution multiplications in the frequency domain increases from $\prod_i N_i$ to $\sum_i \prod_j \frac{N_j}{K_i} \leq m \prod_i N_i$.

% In contrast, optimal non-separable scattering transforms can achieve a computational cost of $\sum_{i} N_i \log N_i + P \log P$, since a full forward \ac{fft} is required to compute a convolution with a non-separable $m$-dimensional filter.

% The greatest computational advantage is during direct convolution computations, as is present in \acp{cnn}. For an $m$-dimensional filter impulse response with filter lengths $L_1, ..., L_m$, the computational cost reduces from $P\prod_i L_i$ to $P \sum_i L_i$.

% We benchmark separable \ac{ws} against 2D scattering by repeating the MNIST hand-written digit dataset experiment in \citep{2dscattering}.

The MNIST dataset \citep{mnist} contains 60000 training and 10000 test samples. Unlike in \citep{2dscattering}, which decorrelates scattering coefficients with a discrete cosine transform prior to classification, we perform classification on the scattering coefficients directly. Kymatio \citep{kymatio} is used to produce the 2D scattering coefficients. Our implementation of separable scattering is implemented similarly to Kymatio, with PyTorch \citep{pytorch} as a backend for FFT convolutions and \ac{nn} models.

\begin{table}
\centering
\caption{MNIST classification error rate (\%) of separable and \ac{2d} scattering coefficients using a \ac{nn} classifier} \label{tab:mnistnn}
\begin{tabular}{|lr|l|} \hline
\multirow{2}{*}{2D WS + NN}        & $l=1$, $J=2$  & $0.64 \pm 0.05$    \\
          & $l=2$, $J=3$ & $0.50 \pm 0.03$     \\ \hline
\multirow{2}{*}{Separable WS + NN}  & $l=1$, $d=(4,4)$ & $0.63 \pm 0.05 $    \\
          & $l=2$, $d=(4,4)$  & $0.52 \pm 0.04$     \\ \hline
\end{tabular}
\end{table}

% Table \ref{tab:mnistlda} shows the results for MNIST using a \ac{lda} classifier from the Scikit-learn python package \citep{sklearn}, with a covariance shrinkage factor of $5\cdot 10^{-3}$. Scattering is performed for $l=1, 2$ levels.

% \begin{table}[!h] 
% \centering
% \caption{MNIST classification error rate (\%) of separable and \ac{2d} scattering coefficients  using an \ac{lda} classifier} \label{tab:mnistlda}
% \begin{tabular}{|r|rr|rr|} \hline
% \multirow{2}{*}{Training size}              & \multicolumn{2}{l}{2D}                               & \multicolumn{2}{l|}{Separable}                             \\
%  & \multicolumn{1}{l}{$l=1$} & \multicolumn{1}{l}{$l=2$} & \multicolumn{1}{l}{$l=1$} & \multicolumn{1}{l|}{$l=2$} \\ \hline
% 1000                              & 4.58                        & 2.69                        & 4.87                        & 4.58                        \\
% 2000                              & 3.89                        & 1.62                        & 4.31                        & 2.6                         \\
% 5000                              & 3.54                        & 1.12                        & 3.84                        & 1.7                         \\
% 10000                             & 3.35                        & 1.05                        & 3.54                        & 1.5                         \\
% 20000                             & 3.16                        & 0.87                        & 3.41                        & 1.41                        \\
% 40000                             & 2.92                        & 0.86                        & 3.34                        & 1.25                        \\
% 60000                             & 2.92                        & 0.81                        & 3.27                        & 1.34     \\ \hline                  
% \end{tabular}
% \end{table}

Due to its separability and non-angularly spaced filters, separable scattering does not perform as well compared to 2D scattering when using simple classifiers, such as \ac{lda} \citep{lda}. To illustrate that this performance discrepancy is not of significant consequence in a \ac{nn} setting, we test performance on the full dataset utilising a simple architecture. The neural network architecture used has an input layer with 256 neurons, followed by two hidden layers with 128 and 64 neurons respectively. The output layer has 10 neurons - one for each digit. Input and hidden layers are followed by a batch norm layer \citep{batchnorm} and ReLU activation function \citep{relu}. The output layer is followed by a softmax function. The Adam optimiser \citep{adam} is used with cross-entropy loss, a batch size of 256 and learning rate of $3 \cdot 10^{-5}$. 5000 of the 60000 training samples are reserved for validation and removed from the training set. Training is stopped when validation loss starts to increase. The \ac{nn} model is initialised with random weights, and the experiment is repeated 50 times. Different invariance scales were tested, and the best results are reported in table \ref{tab:mnistnn}. Tests are repeated for $l \in \{1, 2\}$ levels of scattering.

\subsection{MedMNIST3D}

The MedMNIST3D datasets are a subset of the MedMNIST dataset group \citep{medmnist}, where each 3D dataset contains $28\times 28 \times 28$ images with 2, 3 or 11 classes. Train, test and validation data partitions are provided by the authors. All datasets have on the order of 1000 training samples. We compare the baseline \ac{nn} results provided in \citep{medmnist} with separable scattering features combined with a simple \ac{nn} classifier.

We use \ac{nn} classifier with an input layer containing 1024 neurons, followed by two hidden layers with 512 and 256 neurons respectively. Input and hidden layers layers are each followed by a batch norm layer \citep{batchnorm} and a ReLU non-linearity \citep{relu}. For datasets with two classes, the output layer is a single neuron followed by a sigmoid activation function. For datasets with more than two classes, the output layer has a size equal to the number of classes, followed by a softmax activation. Binary cross-entropy loss are used for datasets with two classes, otherwise cross-entropy loss is used. The Adam optimiser \citep{adam} with a learning rate of $1 \cdot 10^{-5}$ is used. All other configuration parameters are identical to the model used for the MNIST dataset.

>>>>

\chapter{Separable Wavelet Scattering}
\label{chap:p3}

\articleinfo{Seperable Wavelet Scattering}{IEEE Signal Processing Letters}{15 May 2024}{doi}

\section*{CRediT}
\textbf{M.W. Rademan}: conceptualisation; methodology; software; formal analysis; writing -- original draft.

\textbf{D.J.J. Verseld}: writing -- review and editing; supervision.

\textbf{J.A. Du Preez}: writing -- review and editing; supervision.

\section*{Abstract}
Wavelet scattering is a widely used feature extraction method due to it efficacy in extracting invariant features. It retains lost high frequency information resulting from the averaging operations required to obtain invariance. Different wavelet scattering transforms exist for a varying number of input dimensions, each having different choices of wavelets and filterbank construction techniques. In this work, we propose a new scattering transform by generalising the one-dimensional case to an arbitrary number of independent dimensions. Additionally, we provide concise implementation definitions for the Morlet wavelet. We show that, in a modern deep learning setting, separable wavelet scattering performs similarly to its non-separable counterparts with the MNIST hand-written digit dataset. We additionally demonstrate state-of-the-art results for a subset of the MedMNIST3D datasets.

\section{Introduction}
\Ac{ws} has seen widespread use in classification applications as a powerful feature extraction method. It is an especially effective method for small datasets, since the feature filters are reminiscent of a \ac{cnn} front-end, while requiring no learning. The resulting features are invariant across all averaged dimensions, and exhibit separable class subspaces, allowing linear classifiers to be used with high efficacy \cite{2dscattering}.

Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \cite{1dscattering1, ws}, which was later extended to 2 \cite{2dscattering} and 3 dimensions \cite{3dscattering, harmonicscattering}. Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \cite{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \cite{ws_joint_tf, jointtfscattering2}. Joint-\ac{tf} scattering utilises a separable \ac{2d} filter that operates on the first level scattering scalogram, denoted by the operator $\mathcal{U}_1$, which is the only apparent usage of separable filters in the current scattering literature.

In a deep-learning setting, sensible initialisation of filters prior to learning can significantly improve performance and interpretability \cite{sincnet}. The greatest advantage of separable filters is their computation speed when computing convolutions directly, as is performed in \acp{cnn} \cite{separablecnn}. Additionally, separable filters reduce the number of parameters of a \ac{cnn} if the filters are configured to be learnable. Learnable filters can typically improve performance compared to their fixed-filters counterpart \cite{scattering_birdsong}.

At first glance, a \ac{ws} decomposition seems to be prohibitively expensive, but fast algorithms are possible due to the limited bandwidth of the filter output. Fast algorithms utilise downsampling to take advantage of the demodulation of band-limited signals that result from the convolution of analytic wavelets \cite{2dscattering, 3dscattering, 1dscattering1}. Path pruning is also used to ignore filter combinations which have negligible energy. These algorithms are the standard implementation on many platforms, such as the Kymatio python package \cite{kymatio} and MATLAB \cite{MATLAB}.

@@

\subsection{Morlet Wavelet}

A \ac{1d} zero-mean Gaussian is given by
\begin{equation}
    \theta_{\sigma_t}(t) = \frac{1}{\sqrt{2\pi\sigma_t^2}}e^{-\frac{1}{2}\left(\frac{t}{\sigma_t}\right)^2},
\end{equation}
which has a Fourier transform transform $\theta_{\sigma_t}(t) \xleftrightarrow{\mathcal{F}} \hat{\theta}_{\sigma_\omega}(\omega)$, where $\sigma_\omega = \frac{1}{\sigma_t}$.

For an arbitrary bandwidth definition, we define the bandwidth-to-$\sigma$ ratio $\beta \in \mathbb{R}^+$, such that the one-sided bandwidth of a zero-mean Gaussian is $\beta \sigma_\omega$. For reference, Kymatio's implementation of 2D scattering indirectly utilises $\beta \approx 2.5$ \cite{kymatio}.

@@

A \ac{1d} Morlet wavelet $\uppsi$ has energy concentrated around 1 rad/s, with most of its energy contained in the interval $\omega \in [1 - \frac{1}{Q}, 1 + \frac{1}{Q}]$. $Q \in \mathbb{R}^+$ is defined as the number of wavelets per octave. The mother wavelet is given as
\begin{equation}
    \uppsi(t) = \theta_{\sigma_t}(t)\left(e^{jt} - \theta_0 \right),
\end{equation}
where $\theta_0 = \frac{\theta_{\sigma_t}(-1)}{\theta_{\sigma_t}(0)}$ to ensure zero mean: $\hat{\uppsi}(0) = 0$.

For reasons that will become apparent in section \ref{sec:filterbank}, we defined a 1D-wavelet dilated by a factor $\lambda$ as 
\begin{equation}
\label{eqn:dilwav}
    \uppsi_\lambda(t) = \begin{cases}
        \uppsi(\lambda t), \ \lambda \neq 0 \\
        \phi_i(t), \ \lambda = 0
    \end{cases},
\end{equation}
where $\phi$ represents the 1D \ac{lpf} utilised in scattering computations for the current dimension ($i$).

An $n$-dimensional separable wavelet may be constructed similarly, with
\begin{equation}
    \uppsi_{\vect{\lambda} }(\mathbf{u}) = \prod_{i=0}^{n-1} \uppsi_{\lambda_i}(u_i),
\end{equation}
where $\mathbf{u} = \left(u_1, ..., u_n\right)^T$ is the vector containing the dimensions of interest.

@@

\subsection{1D Filterbank Construction}
\label{sec:filterbank}
 Suppose a 1D Morlet wavelet has a centre frequency $\lambda_0$. The following filter is placed at $\lambda_0 2^{\frac{1}{Q}}$. To define the amount of overlap between filters, it is useful to express the corresponding filter's frequency \ac{std} $\sigma_\omega$ in terms of the distance between filters in the frequency domain. Suppose a wavelet at centre frequency $\lambda_0$ requires a decay equivalent to $\alpha \sigma_\omega$ \ac{std}s at the next wavelet at $\lambda_0 2^{\frac{1}{Q}}$. It follows that $\sigma_\omega = \frac{1}{\sigma_t} = \frac{1}{\alpha}\left( 2^\frac{1}{Q} - 1 \right)$.

@@

The provided definitions of filter clearance ($\alpha$) and bandwidth ($\beta$) may also be extended to non-separable wavelets, such as the rotational construction employed in \cite{2dscattering}. Other scattering transforms may also benefit from the downsampling strategies proposed in this paper.

@@

The scattering transform requires that the time support of all filters do not exceed the time support of $\phi$, ensuring no filters contain time information exceeding the invariance scale $d$. The invariance scale $d$ is equivalently defined as the total downsample factor of the scattering tranform, and is unique for each dimension. Equivalently, the frequency \ac{std} of the filters may not exceed the \ac{std} of $\phi(\omega)$. For brevity, we denote the frequency \ac{std}s as $\sigma_{\lambda, \omega}$ and $\sigma_{\phi, \omega}$ for the Morlets and the \ac{lpf} respectively.

@@

Only positive $\lambda$'s have been defined thus far, which provides inadequate coverage of the frequency domain in multiple dimensions. For real input signals, it is only necessary to cover half of one of the dimensions (only positive $\lambda$'s), whereas full coverage (both negative and positive $\lambda$'s) is required for additional dimensions. On-axis coverage is also required, in which each $\uppsi_\lambda$ must be multiplied with a Gaussian (zero-frequency wavelet), which the dilated wavelet definition in equation (\ref{eqn:dilwav}) defines as $\lambda=0$. A similar construction procedure is followed in \cite{jointtfscattering2}.

Given $m$ 1D filterbanks, with $m \ge 2$, each having Morlet filters with a positive set of centre frequencies $\Lambda_i^+$ and invariance scales $\vect{d} \in \mathbb{N}^m$, with $i = 1, ..., m$ indexing the dimension, we construct the $m$-dimensional filterbank with 
\begin{equation}
    \mathbb{F} = \left\{ \uppsi_{\vect{\lambda} }(\mathbf{u}) \ \middle| \ \vect{\lambda} \in \mathbb{L} \setminus \{\vect{0}\} \right\},
\end{equation}
where $\vect{u}$ is the $m$-dimensional spatial and/or time variable in which the each 1D filterbank is defined.  By definition, 
\begin{gather}
    \mathbb{L} = (\Lambda_1^+ \!\cup\! \{0\})\!\times\! (\Lambda_2^+ \!\cup\! \Lambda_2^- \!\cup \!\{0\}) \!\times\! ... \!\times\! (\Lambda_m^+ \!\cup\! \Lambda_m^- \!\cup\! \{0\}); \\
    \Lambda_i^- = \left\{-\lambda \ \middle| \ \lambda \in \Lambda_i^+\right\},
\end{gather}
where $\cup$ indicates the set union operator and $\times$ the Cartesian product. For $m=1$, the provided definitions result in a conventional \ac{1d} scattering transform \cite{1dscattering1}.

@@

\section{Separable Scattering Transform} \label{sec:wst}
% In this section, we describe the efficient computation of $m$-dimensional separable scattering transforms. The described downsampling strategy is utilised to a limited extent in Kymatio scattering implementations \cite{kymatio} and the original proposed fast algorithms \cite{2dscattering}, which is generalised further in this section.

@@

A downsample factor $d$ decomposes into a set of $n$ prime factors $\{p_1, ..., p_n\}$ with a corresponding multiset $\mathbb{M} = \left\{ m_1, ..., m_n \right\}$, where $m_i$ is the multiplicity of the prime $p_i$. We can find an optimal $d$ by maximising the sum $\sum\limits_{m \in \mathbb{M}} m$.

@@

We can efficiently compute $s_1$ using compounded downsampling steps:
\begin{equation}
    \left(s_1\right)_{\downarrow d} = \left( \left(u_1\right)_{\downarrow d_1} * (\phi)_{\downarrow d_1}\right)_{\downarrow d_2},
\end{equation}
such that $d = d_1 \cdot d_2, \ d_1, d_2 \in \mathbb{N}^+$, with $d_1 | d$ and $d_1 \le  d_{\uppsi_1}$. The notation $x | y$ indicates that $x$ evenly divides $y$. In order to find $d_1$, we decrement $d_{\uppsi_1}$ until it divides $d$ evenly. Each scalogram $u_j[n]$ is not necessarily downsampled optimally, but has a downsampling factor which guarentees a consistent scattering coefficient output sample frequency.

@@

Given that $r|N$, we can express (\ref{eqn:circconv}) when downsampled as a periodised summation \cite{waveletsandsubbandcoding} in the frequency domain
\begin{equation}
    (x \otimes \uppsi [n])_{\downarrow r} \xleftrightarrow{\mathcal{FFT}}  \frac{1}{r} \sum_{i=0}^{r-1} \hat{x} \cdot \hat{\uppsi} [k + iN/r], %https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=5216ea733e562541b33a7f97dab0de072b2e8827
\end{equation}
which can be efficiently implemented via shape manipulation of tensors in computational packages like MATLAB or PyTorch \cite{pytorch}. To compute valid convolutions, we must pad $x$ and $\uppsi$ to have a total length of $N = N_x + d + c$, where $c \in \mathbb{N}^+$ is a constant that ensures that $d | N$.

% We benchmark separable \ac{ws} against 2D scattering by repeating the MNIST hand-written digit dataset experiment in \cite{2dscattering}.

The MNIST dataset \cite{mnist} contains 60000 training and 10000 test samples. Unlike in \cite{2dscattering}, which decorrelates scattering coefficients with a discrete cosine transform prior to classification, we perform classification on the scattering coefficients directly. Kymatio \cite{kymatio} is used to produce the 2D scattering coefficients. Our implementation of separable scattering is implemented similarly to Kymatio, with PyTorch \cite{pytorch} as a backend for FFT convolutions and \ac{nn} models.

\begin{table}
\centering
\caption{MNIST classification error rate (\%) of separable and \ac{2d} scattering coefficients using a NN classifier.} \label{tab:mnistnn}
\begin{tabular}{|lr|l|} \hline
\multirow{2}{*}{2D WS + NN}        & $l=1$, $J=2$  & $0.64 \pm 0.05$    \\
          & $l=2$, $J=3$ & $0.50 \pm 0.03$     \\ \hline
\multirow{2}{*}{Separable WS + NN}  & $l=1$, $d=(4,4)$ & $0.63 \pm 0.05 $    \\
          & $l=2$, $d=(4,4)$  & $0.52 \pm 0.04$     \\ \hline
\end{tabular}
\end{table}

Due to its separability and non-angularly spaced filters, separable scattering does not perform as well compared to 2D scattering when using simple classifiers, such as \ac{lda} \cite{lda}. To illustrate that this performance discrepancy is not of significant consequence in a \ac{nn} setting, we test performance on the full dataset utilising a simple architecture. The neural network architecture used has an input layer with 256 neurons, followed by two hidden layers with 128 and 64 neurons respectively. The output layer has 10 neurons - one for each digit. Input and hidden layers are followed by a batch norm layer \cite{batchnorm} and ReLU activation function \cite{relu}. The output layer is followed by a softmax function. The Adam optimiser \cite{adam} is used with cross-entropy loss, a batch size of 256 and learning rate of $3 \cdot 10^{-5}$. 5000 of the 60000 training samples are reserved for validation and removed from the training set. Training is stopped when validation loss starts to increase. The \ac{nn} model is initialised with random weights, and the experiment is repeated 50 times. Different invariance scales were tested, and the best results are reported in table \ref{tab:mnistnn}. Tests are repeated for $l \in \{1, 2\}$ levels of scattering.

\subsection{MedMNIST3D}
The MedMNIST3D datasets are a subset of the MedMNIST dataset group \cite{medmnist}, where each 3D dataset contains $28\times 28 \times 28$ images with 2, 3 or 11 classes. Train, test and validation data partitions are provided by the authors. All datasets have on the order of 1000 training samples. We compare the baseline \ac{nn} results provided in \cite{medmnist} with separable scattering features combined with a simple \ac{nn} classifier.

We use \ac{nn} classifier with an input layer containing 1024 neurons, followed by two hidden layers with 512 and 256 neurons respectively. Input and hidden layers layers are each followed by a batch norm layer \cite{batchnorm} and a ReLU non-linearity \cite{relu}. For datasets with two classes, the output layer is a single neuron followed by a sigmoid activation function. For datasets with more than two classes, the output layer has a size equal to the number of classes, followed by a softmax activation. Binary cross-entropy loss are used for datasets with two classes, otherwise cross-entropy loss is used. The Adam optimiser \cite{adam} with a learning rate of $1 \cdot 10^{-5}$ is used. All other configuration parameters are identical to the model used for the MNIST dataset.

A single level of scattering coefficients are computed, with $Q = (2, 2, 2)$ and $d = (4, 4, 4)$. Many MedMNIST3D datasets tend to be unbalanced, implying that \ac{auc} is a more reliable metric to measure model performance. The results are shown in table \ref{tab:medmnist}, in which \ac{sota} \ac{auc} performance is achieved for the Organ, Adrenal and Vessel datasets. Table \ref{tab:medmnist} indicates the number of classes ($c$) for each of the datasets, with accuracy (ACC) also shown for reference. Non-\ac{sota} results are comparable to the other baseline \ac{nn} approaches presented in \cite{medmnist}. It is likely that better results can be achieved by the proposed method if the filters are made learnable and/or scattering parameters and network architectures are uniquely optimised for each dataset.

\begin{table*}[]

@@

    \caption[MedMNIST classification results of compared to baseline \ac{nn} approaches.]{MedMNIST classification results of compared to baseline NN approaches (\citep{medmnist}).} \label{tab:medmnist}
    \begin{adjustbox}{angle=90}

@@

\end{adjustbox}

@@
